Explaining approximate deterministic inference methods

### 2022-11-30 src/docs_xx-12-22-explainer/xx-12-22-explainer.Rmd
Working on lab group presentation

* As a statistical modeller, our whole job (approximately) is to construct a generative model for data $y$ using parameters $\vartheta$
* This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**
$$
p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}
$$
* The **core problem** of Bayesian inference is that we can't compute $p(y)$
$$
p(y) = \int p(y, \vartheta) \text{d} \vartheta
$$

@@

* Usually if you wanted to integrate something, you could use numerical integration methods
* It doesn't work when $\text{dim}(\vartheta)$ gets even moderately sized

# Notation set-up

* Previously I had all of the parameters under the symbol $\vartheta$
* What if we split them up as being $\vartheta = (x, \theta)$

>>>>

* As a statistical modeller, our whole job (approximately) is to construct a generative model for data $y$ using parameters $\vartheta$
* This is the joint distribution $p(y, \vartheta) = p(y \, | \, \vartheta) p(\vartheta)$
* What we actually want is the posterior $p(\vartheta \, | \, y)$ which is **just**^[I've bolded this for sarcasm.]
$$
p(\vartheta \, | \, y) = \frac{p(y, \vartheta)}{p(y)} = \frac{p(y \, | \, \vartheta) p(\vartheta)}{p(y)}
$$
* The **core problem** of Bayesian inference is that we can't compute $p(y)$
$$
p(y) = \int p(y, \vartheta) \text{d} \vartheta
$$

@@

* Usually if you wanted to integrate something, you could use numerical integration methods
* Pick nodes $\vartheta_i$ and weights $\omega_i$ then compute the sum
$$
\tilde p(y) = \sum_i p(y, \vartheta_i) \omega_i
$$

# Monte Carlo as an example of numerical integration

* Suppose we can sample $\vartheta_i \sim p(y, \vartheta)$ for $i = 1, \ldots, N$
* If we set $\omega_i = 1/N$ for all $i$ then
$$
\tilde p(y) = \frac{1}{N} \sum_i p(y, \vartheta_i)
$$

# Monte Carlo is fundamentally unsound

* "Monte Carlo ignores information" according to @o1987monte
* Suppose $N = 3$ and we sample $\vartheta_1, \vartheta_2, \vartheta_3$ with $\vartheta_2 = \vartheta_3$ then our MC estimate is
$$
\tilde p(y) = \frac{1}{3} \left( p(y, \vartheta_1) + p(y, \vartheta_2) + p(y, \vartheta_3) \right)
$$
* This is despite the fact that nothing new about the function has been learned by adding $\{\vartheta_3, p(y, \vartheta_3)\}$

# Application to HIV survey sampling

* Suppose we're running a household survey, and sample the same individual twice
* We didn't learn anything new about HIV by surveying them again!
* This doesn't just bite for nodes or individuals which are exactly the same: an analogous argument can be made if they are close together and we expect their function evaluations to be similar

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
$\implies$ Bayesian quadrature, Bayesian survey design (end of digression)
\end{tcolorbox}
\end{center}

# Latent variables and hyperparameters

* Quadrature doesn't work very well when $\text{dim}(\vartheta)$ gets even moderately sized
* Previously I had all of the parameters under the symbol $\vartheta$
* What if we split them up as being $\vartheta = (x, \theta)$
* The key part about this is that $\dim(x)$ is big and $\dim(\theta)$ is small
